```{r, setoptions,echo=FALSE, results='hide'}
library(knitr)
library(gridExtra)
library(ggplot2)
opts_chunk$set(echo = TRUE)
```
Impact of Hydro-Meteorological Events
=====================================
## US Storm Data, 1950-2011


### Synopsis:  
The impact of a storm can range from insignificant to immense. In this document we attempt to take a cursory glance through what might be the most impactful storm types according to all the data the NOAA has to offer from 1950 through 2011.
  

### Data Processing:
To begin with, our data comes in the form of a rather large Comma Separated Value file. Roughly half a gigabyte, this file contains just over 900 thousand observations, spaning more than 60 years. The data for the last 20 or 30 years is considerably, and understandably, more complete than the years prior. This data is available from the [NOAA Storm Events Database](http://www.ncdc.noaa.gov/stormevents/ftp.jsp) in CSV files separated by year. Our dataset is presented as a conglomerate of all the data from the years mentioned above.  

```{r download.data, results='hide', echo=FALSE}
if (!file.exists('stormdata.csv')) {
    # attempt to download the file.
    file.url = 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
    if (.Platform$OS.type == 'unix') {
        # Download the file using CURL
        download.file(file.url, 'stormdata.csv.bz2', method='curl')
    } else {
        # Download the file using internal system methods
        download.file(file.url, 'stormdata.csv.bz2')
    }
    # Use R.utils package to unzip file.
    bunzip2('stormdata.csv.bz2')
}
```

To begin with we will need to take the data from it's raw text form and bring it into R
to start our analysis. We will be using the `data.table` package to make our processing
faster after our initial read.  
You'll note I'm using the `base` packages' `read.csv()` function and then immediately
turning that `data.frame` into a `data.table`. This is because there is a known and
tracked bug when using `fread` to read .csv files that have irregular usage of commas
inside escaped blocks of text. That being said, the method below only takes a fractional
amount of time more and will yeild considerably faster processing below:  
```{r read.data, cache=TRUE, results='hold'}
library(data.table)
data <- data.table(read.csv('stormdata.csv'))
```

Next, we will begin the actual processing of our data. First, we are going to convert
the data that corresponds with the fiscal impact of an event into whole numbers.
Having seen the original data, there seems to be some mis-translation of the original
numbers into the form our current csv file has. We will be doing our best to accomodate
and take only what we know for certain at face value:  
```{r process.dmg, cache=TRUE, results='hide', dependson=c('read.data')}
# First we'll give ourselves something like a row number to work our data with
data[,r:=1:.N]

# Each of the blocks below follow the same functionality.
# Loop through all values of a given variable (PROPDMG or CROPDMG), and
# use the data in the corresponding "*EXP" variable to convert that into
# a whole number. If a number is NA, we are going to record it as 0, this has
# not been found to skew our results at all. If the exponent recorded is not
# one of a clearly defined few (this is where that uncertainty I mentioned  from
# the original data sources comes in) we are going to just take the value as it is.
data[,PROPDMG:={
    if (PROPDMGEXP == 'b' | PROPDMGEXP =='B') {
        PROPDMG*1000000000
    } else if (PROPDMGEXP == 'm' | PROPDMGEXP =='M') {
        PROPDMG*1000000
    } else if (PROPDMGEXP == 'k' | PROPDMGEXP =='K') {
        PROPDMG*1000
    } else if (PROPDMGEXP == 'h' | PROPDMGEXP == 'H') {
        PROPDMG*100
    } else if (is.na(PROPDMG)) {
        0
    } else {
        PROPDMG
    }
},by=r]

data[,CROPDMG:={
    if (CROPDMGEXP == 'b' | CROPDMGEXP =='B') {
        CROPDMG*1000000000
    } else if (PROPDMGEXP == 'm' | CROPDMGEXP =='M') {
        CROPDMG*1000000
    } else if (PROPDMGEXP == 'k' | CROPDMGEXP =='K') {
        CROPDMG*1000
    } else if (PROPDMGEXP == 'h' | PROPDMGEXP == 'H') {
        PROPDMG*100
    } else if (is.na(CROPDMG)) {
        0
    } else {
        CROPDMG
    }
},by=r]
```


This next step summarizes all of the data for our calculations. It will take a total and average for the statistics we are interested in: Damage as measured by Cost (USD), Fatalities, and Injuries.
```{r summarize.data, cache=TRUE, dependson=c('read.data', 'process.dmg')}
#cost
# Calculate total and average data for each of the following statistics
# Damage (in dollars), Injuries (# people), Fatalities (# people)
data.by.type <- data[,
                     list(
                         sum(PROPDMG+CROPDMG),
                         round(mean(PROPDMG+CROPDMG),digits=0),
                         sum(INJURIES),
                         round(mean(INJURIES), digits=1),
                         sum(FATALITIES),
                         round(mean(FATALITIES), digits=1),
                         .N),
                     by=EVTYPE]
setnames(data.by.type, c('EVTYPE','SUMDMG','AVGDMG','SUMINJ', 'AVGINJ', 'SUMFAT', 'AVGFAT', 'COUNT'))
# Display a small representation of the table we've built
data.by.type
```

# Results of the Analysis
### Damage as measured by Cost (USD)

Below you'll see the code used to generate a Figure that details the 5 most Costly weather events from our dataset. I've shown 2 plots here, one that shows the highest by **total** and one that shows the highest by **average**. This shows that there is a distinct difference between these two figures.
```{r plot.cost, cache=TRUE, dependson=c('summarize.data'), fig.height=10, fig.width=10}
# Order the data by total damage
data.by.type <- data.by.type[order(-SUMDMG),]
# Plot the 5 highest costing event types, by total cost, as a horizontal bar plot
# This first figure also includes point data to show the huge range of outliers versus
# the relatively small range of the bulk of the population.
total.figure <- ggplot(data=data.by.type[5:1,],aes(EVTYPE, SUMDMG, fill=EVTYPE)) +
    # define this as a bar plot of identities (ie not a histogram)
    geom_bar(stat="identity") +
    #modify the x and y scales to show our data more clearly
    scale_x_discrete(limits=data.by.type[5:1,EVTYPE]) +
    scale_y_continuous(breaks=c(0,5e+10,1e+11,1.5e+11),labels=c(0,50,100,150)) +
    #label each bar with the corresponding event type and total cost
    geom_text(aes(EVTYPE,rep_len(min(data.by.type$SUMDMG),5),
                  label=paste(EVTYPE, '-', '$', format(SUMDMG/1e+09, digits=3), 'B', sep=' ')),
              hjust=0, vjust=2) +
    #graph individual data to show the grouped population versus the outliers
    geom_point(data=data[EVTYPE %in% data.by.type[5:1,EVTYPE]],
               aes(EVTYPE, (PROPDMG+CROPDMG), fill=EVTYPE), shape=23,
               alpha=0.75, position=position_jitter(height=.5, width=0)) +    
    # label the plot, color and fill bars
    labs(title='Highest Total Costs') +
    theme_bw() +
    theme(plot.title=element_text(vjust=1, face='bold', size='16'),
          axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank()) + 
    guides(fill=F) +
    # flip the x and y coordinates to make this plot horizontal.
    coord_flip()

# Same plot as above, but uses averages instead of totals.
data.by.type <- data.by.type[order(-AVGDMG),]
avg.figure <- ggplot(data=data.by.type[5:1,],aes(EVTYPE, AVGDMG, fill=EVTYPE)) +
    geom_bar(stat="identity") +
    scale_x_discrete(limits=data.by.type[5:1,EVTYPE]) +
    scale_y_continuous(breaks=c(0,5e+08,1e+09,1.5e+09),labels=c(0,0.5,1.0,1.5)) +
    geom_text(aes(EVTYPE,rep_len(min(data.by.type$AVGDMG),5),
                  label=paste(EVTYPE, '-', '$', format(AVGDMG/1e+09, digits=3), 'B', sep=' ')),
              hjust=0, vjust=2) +
    labs(title='Highest Average Costs') +
    theme_bw() +
    theme(plot.title=element_text(vjust=1, face='bold', size='16'),
          axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank()) + 
    guides(fill=F) +
    coord_flip()
# Use gridExtra library to display these two charts as one figure.
grid.arrange(total.figure,
             avg.figure,
             ncol=1,
             main=textGrob("Most Costly Weather Event Types - 1950 to 2011",
                           gp=gpar(fontsize=20, fontface="bold"), just='top')
)
```

You can see from these plots that there is a HUGE difference between the totals and the averages. We can see from the first chart that the total damage caused by Flooding since 1950 is *considerably* higher than the next weather event type on the list. However, when we look at the second chart we can see that "Flood" isn't even on the top five when averaged out per event.  

So which type of Weather Event should you prepare for? Regardles of the specific types it seems that **water damage** is a common factor both in the totals chart and the averages chart. Flooding can often be a secondary effect of other weather event types. The only common specific type between these two charts is **Hurricane/Typhoon** - a weather event that generates huge amount of precipitation and often causes flooding. I would put a large portion of any Disaster Planning budget into mitigating the devestating effects of water on agricultural and residential property.

### Damage as measured by Fatalities and Injuries  

We will once again see these statistics as totals and averages separately to illustrate the differences.
```{r plot.health, cache=TRUE, dependson=c('summarize.data'), fig.height=12, fig.width=12}
# Total Injuries
data.by.type <- data.by.type[order(-SUMINJ),]
total.inj <- ggplot(data=data.by.type[5:1,],aes(EVTYPE, SUMINJ, fill=EVTYPE)) +
    geom_bar(stat="identity") +
    scale_x_discrete(limits=data.by.type[5:1,EVTYPE]) +
    geom_text(aes(EVTYPE,rep_len(0,5),label=paste(EVTYPE, SUMINJ, sep=' - ')),
              hjust=0, vjust=2) +
    labs(title='Highest Total Injuries') +
    theme_bw() +
    theme(plot.title=element_text(vjust=1, face='bold', size='16'),
          axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank()) + 
    guides(fill=F) +
    coord_flip()

# Average Injuries
data.by.type <- data.by.type[order(-AVGINJ),]
avg.inj <- ggplot(data=data.by.type[5:1,],aes(EVTYPE, AVGINJ, fill=EVTYPE)) +
    geom_bar(stat="identity") +
    scale_x_discrete(limits=data.by.type[5:1,EVTYPE]) +
    scale_y_continuous(breaks=c(0,5e+08,1e+09,1.5e+09),labels=c(0,0.5,1.0,1.5)) +
    geom_text(aes(EVTYPE,rep_len(0,5),label=paste(EVTYPE, AVGINJ, sep=' - ')),
              hjust=0, vjust=2) +
    labs(title='Highest Average Injuries') +
    theme_bw() +
    theme(plot.title=element_text(vjust=1, face='bold', size='16'),
          axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank()) + 
    guides(fill=F) +
    coord_flip()

# Total fatalities
data.by.type <- data.by.type[order(-SUMFAT),]
total.fat <- ggplot(data=data.by.type[5:1,],aes(EVTYPE, SUMFAT, fill=EVTYPE)) +
    geom_bar(stat="identity") +
    scale_x_discrete(limits=data.by.type[5:1,EVTYPE]) +
    geom_text(aes(EVTYPE,rep_len(0,5),label=paste(EVTYPE, SUMFAT, sep=' - ')),
              hjust=0, vjust=2) +
    labs(title='Highest Total Fatalities') +
    theme_bw() +
    theme(plot.title=element_text(vjust=1, face='bold', size='16'),
          axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank()) + 
    guides(fill=F) +
    coord_flip()

# Average Fatalities
data.by.type <- data.by.type[order(-AVGFAT),]
avg.fat <- ggplot(data=data.by.type[5:1,],aes(EVTYPE, AVGFAT, fill=EVTYPE)) +
    geom_bar(stat="identity") +
    scale_x_discrete(limits=data.by.type[5:1,EVTYPE]) +
    scale_y_continuous(breaks=c(0,5e+08,1e+09,1.5e+09),labels=c(0,0.5,1.0,1.5)) +
    geom_text(aes(EVTYPE,rep_len(0,5),label=paste(EVTYPE, AVGFAT, sep=' - ')),
              hjust=0, vjust=2) +
    labs(title='Highest Average Fatalities') +
    theme_bw() +
    theme(plot.title=element_text(vjust=1, face='bold', size='16'),
          axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank()) + 
    guides(fill=F) +
    coord_flip()

# Plot all four charts in a figure
grid.arrange(total.inj,
             avg.inj,
             total.fat,
             avg.fat,
             ncol=2,
             main=textGrob("Most Dangerous Weather Event Types - 1950 to 2011",
                           gp=gpar(fontsize=20, fontface="bold"), just='top')
)
```

Here, again, you can see that there are huge discrepencies between a look into the **total** harm done and the **average** harm done as measured separately in fatalities and non-fatal injuries. These charts, however, show some crossover between them. **Tornadoes** and related weather events are represented in some way on every chart, as are **heat** related injuries. In terms of the human cost to weather it would seem that injury most often arises from wind-swept debris or from extreme elements such as heat or cold.